---
title: "COVID-19 time series analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(fs)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(tsibble)
library(feasts)
```

## Initialise

```{r init}
case_data_url <- "https://github.com/minhealthnz/nz-covid-data/blob/main/cases/covid-cases.csv?raw=true"
dhb_name <- "Capital and Coast"
```

## Download latest data

```{r download}
cases_df <- read_csv(case_data_url) %>% 
  clean_names()

head(cases_df)

skim(cases_df)

```

## DHB summaries

```{r dhb-summary}

cases_by_dhb_df <- cases_df %>% 
  count(report_date, dhb, case_status)

cases_by_dhb_df

```

## Wellington figures

It might be good to add Hutt to CCDHB later.

```{r wellington}
dhb_cases_df <- cases_by_dhb_df %>% 
  filter(dhb == dhb_name) %>% 
  filter(report_date >= ymd(20220201)) %>% 
  group_by(report_date, dhb) %>% 
  summarise(cases = sum(n, na.rm = FALSE))

wfh_start_date <- ymd(20220221)

dhb_cases_df %>%
  ggplot() +
  geom_vline(xintercept = wfh_start_date, colour = "firebrick") +
  geom_line(aes(x = report_date, y = cases)) +
  geom_point(aes(x = report_date, y = cases)) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(date_breaks = "1 week", date_labels = "%d %b") +
  labs(x = "", y= "", title = str_glue("Daily COVID-19 Cases for {dhb_name} DHB")) +
  theme_minimal()
  
```

## Basic timeseries analysis

To best capture the weekly seasonality, it might be best to start modelling from about the time that the reporting process was changed by RATs: probably about the beginning of March.

```{r basic-ts}

# model_start_date <- ymd(20220101)
model_start_date <- ymd(20220301)

dhb_cases_ts <- dhb_cases_df %>% 
  ungroup() %>% 
  filter(report_date >= model_start_date) %>% 
  select(report_date, cases) %>% 
  as_tsibble(index = report_date) %>% 
  fill_gaps(cases = 0) %>% 
  replace_na(list(cases = 0))

dhb_cases_stl <- dhb_cases_ts %>% 
  model(STL(cases ~ trend(window = 7 * 3) + season(period = "1 week", window = Inf))) %>% 
  components()

dhb_cases_stl %>% 
  autoplot()

```

```{r seasonally-adjusted}

dhb_cases_stl %>% 
  select(report_date, cases, trend, season_adjust) %>% 
  ggplot() +
  # geom_vline(xintercept = wfh_start_date, colour = "firebrick") +
  geom_line(aes(x = report_date, y = season_adjust), colour = "firebrick", size = 1) +
  geom_line(aes(x = report_date, y = trend), colour = "steelblue", size = 1) +
  # geom_line(aes(x = report_date, y = cases)) +
  geom_point(aes(x = report_date, y = cases)) +
  scale_x_date(date_breaks = "1 week", date_labels = "%d %b") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "", y= "", title = str_glue("Daily COVID-19 Cases for {dhb_name} DHB")) +
  theme_minimal()

```

For more rigourous analysis, and before we try any forecasting, we should do a log + 1 transform, to allow for multiplicative seasonality.